---
title: "Assignment 6 - PCA"
author: "Mariano Niño Oliva"
date: "1/21/2022"
output: 
  html_document:
    keep_md: true
---

# Packages

```{r}
library("tidyverse")
library("FactoMineR")
library("ggthemes")
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Assignment 6 - PCA on Pokemon Dataset

```{r cars}
getwd()
pokemon <- read.csv("Pokemon.csv")
colnames(pokemon)
```

We extract the needed variables for analysis.
```{r}
poke <- pokemon %>%
  select(HP, Attack, Defense, Sp_Atk, Sp_Def, Speed)

head(poke)
```


Since the units of measurement for the variables are not uniform, we perform PCA with standardization

##Using *prcomp* function

```{r}
pc_poke1 <- prcomp(x = poke, scale. = T)
summary(pc_poke1)

pc_poke1$sdev
pc_poke1$rotation
```

We use another function for comparison

# Using *princomp* function
```{r}
pc_poke2 <- princomp(x = poke, cor = T)
summary(pc_poke2)

```

# Using *PCA* function
## With standardizing
```{r}
pc_poke3 <- PCA(X = poke, scale.unit = TRUE, ncp = 6)
summary(pc_poke3)
pc_poke3$eig
```

**Scree Plot**
```{r}
tibble(eigenvalues = pc_poke3$eig[,1], PC = 1:6) %>%
  ggplot(aes(y = eigenvalues, x = PC)) +
  geom_point() +
  geom_line()
```

We compute the PC scores for the pokemons


```{r}
pc_scores <- as.data.frame(pc_poke2$scores)
head(pc_scores)
with_scores <- bind_cols(pokemon, pc_scores)
with_scores
```

## PC1 vs. PC2 

```{r}
with_scores %>%
  mutate(id = row_number()) %>%
  ggplot(aes(x = Comp.1, y = Comp.2)) +
  geom_text(aes(label = id), col = "red") +
  geom_point()
```

```{r}
with_scores[231,] %>%
  select(HP, Attack, Defense, Sp_Atk, Sp_Def, Speed)
with_scores[430,] %>%
  select(HP, Attack, Defense, Sp_Atk, Sp_Def, Speed)
with_scores[435,] %>%
  select(HP, Attack, Defense, Sp_Atk, Sp_Def, Speed)
```
